---
title: "Final Exam"
author: "Tim Boudreau"
date: "March 16 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache=TRUE)
```

```{r package-load, message = FALSE, warning = FALSE}
library(nycflights13)
library(RSQLite)
library(dplyr)
library(ggplot2)
library(class)
```

## Use the NYC Flight data from 2013:

I split the 2013 NYC Flight Data into five data frames, one for each of the relevant categoried - flights, weather, airlines, airports and planes.
```{r, echo = F, message = FALSE, results = "hide"}
my.data <- nycflights13_sqlite()

flights <- tbl(my.data, "flights") %>% 
  collect() %>%
  mutate(canceled = is.na(arr_time))
flights$canceled <- flights$canceled + 0

weather <- tbl(my.data, "weather") %>% collect()
airlines <- tbl(my.data, "airlines") %>% collect()
airports <- tbl(my.data, "airports") %>% collect()
planes <- tbl(my.data, "planes") %>% collect()

normalize <- function(x) {
  num <- x - min(x, na.rm = T)
  denom <- max(x, na.rm = T) - min(x, na.rm = T)
  return (num/denom)
}
```

I then created data frames for the following: weather data and flight delays and cancelations, airport and flight delays and cancelations, plane characteristics and flight delays and cancelations.
```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
###############
### Weather ###
###############
flights.w <- flights %>%
  group_by(year, month, day, hour) %>%
  summarise(
    delay = mean(dep_delay, na.rm = T),
    cancel = sum(canceled)
  )

weather.w <- weather %>%
  arrange(year, month, day, hour)

data.w <- inner_join(flights.w, weather.w, by = c("year", "month", "day", "hour"))

###############
### Airport ###
###############

flights.ap <- flights %>%
  group_by(year, month, day, hour, dest) %>%
  summarise(
    delay = mean(dep_delay, na.rm = T),
    cancel = sum(canceled)
  ) 

airports.ap <- airports[,1:2]

data.ap <- inner_join(flights.ap, airports.ap, by = c("dest" = "faa"))

#############
### Plane ###
#############

flights.p <- flights %>%
  group_by(year, month, day, hour, tailnum) %>%
  summarise(
    delay = mean(dep_delay, na.rm = T),
    cancel = sum(canceled)
  ) 

data.p <- inner_join(flights.p, planes, by = c("tailnum"))
```

# Data Analysis:

The next section is where I analyze our NYC Flight Data from 2013. Included are figures and statistical significance levels, while regression out put and code is hidden, for cleanliness.

## Delay & Cancelation

First off, we should note that hourly delay has a significant effect, at the 99% confidence level, on the total number of canceled flights per hour.
``` {r echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
summary(glm(cancel~delay, data = data.w))
```

## Weather

No variable has a statistically significant effect on normalized hourly canceled flights.
```{r echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
#### Normalized cancel, probit model ####

summary(glm(normalize(cancel)~temp+dewp+humid+wind_dir+wind_speed+
              wind_gust+precip+pressure+visib, 
            family=binomial(link="probit"),data = data.w))
```

Here, temperature, humidity, pressure and visibility have 99% confident, statistically significant effects on delay time. Wind direction, though, is statistically significant at the 90% level.
```{r echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
#### Hourly delay ####

summary(glm(delay~temp+humid+wind_dir+wind_speed+precip+
              pressure+visib, data = data.w))

```

Temperature, pressure and visibility have a significant effect on the amount of canceled hourly flights, at confidence levels of 99%, 95% and 95% respectively. Wind direction is also significant, though only ata 90% confidence level and when we include other non-significant variables.
```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
#### Hourly canceled flights ####

summary(glm(cancel~temp+humid+wind_dir+wind_speed+precip+pressure
            +visib,data = data.w))

```

As seen in the figure below, a higher temperature leads to more and longer delays. The larger, lighter-colored points are cancelations, which also appear more frequently with higher temperatures.
```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4,fig.cap = "Delay length according to temperature. The color is whether or not the flight was canceled."}
ggplot(data = data.w, aes(x = temp, y = delay, colour = cancel, size = cancel), main = "Temperature and Delays", xlab = "Temperatures", ylab = "Delay Length (minutes)") + scale_x_log10()  + geom_point( alpha = 1) + scale_size(range = c(1,10))
```

## Time

Out of the time variables, month, hour and minute are statistically significant at the 99% confidence level. Day isn't statistically significant.
```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
data.t <- flights %>%
  group_by(year, month, day, hour, minute) %>%
  summarise(
    delay = mean(dep_delay, na.rm=T),
    cancel = sum(canceled)
  )
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
summary(glm(delay~month+day+hour+minute, data = data.t))
```

The variables that significantly impact number of flights canceled are hour, at the 99% level, and minutes, at the 95% level.
```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
summary(glm(cancel~month+day+hour+minute, data = data.t))

```

This graph shows the length of delays relative to the hour of the day the delay occurs. Between midnight and 5 A.M. EST, there are no short delays and, between 1 A.M. and 5 A.M., there are very few longer delays as well. The amount of short delays begins to decrease around 8 P.M., while the amount of long delays begins to increase around 5 A.M.
```{r, echo = FALSE,  message = FALSE, warning = FALSE, fig.height = 4, fig.cap = "Delay length by time of day, where midnight is both hour 0 and hour 24."}
ggplot(data = data.t, aes(x = delay, y = hour, colour = hour), main = "Delay Length by Hour of Day", xlab = "Delay Length (minutes)", ylab = "Hour of Day (Midnight equals 0, 24)") + 
  scale_x_log10()  + geom_point( alpha = 1) + 
  scale_size(range = c(1,10))
```
## Airport

Only 25 of the airport destinations significantly effect the delay length. 
```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
reg.ap <- summary(glm(delay~dest, data = data.ap))

sig <- 0
j <- 1

for (i in 2:(nrow(coef(reg.ap)))) {
  if (coef(reg.ap)[i,4] < .05) {
    sig[j] <- i
    j <- j + 1
  }
}

all.dest <- sort(unique(data.ap$dest))

all.sig.dest <- all.dest[sig]

sig.dest.rows <- match(all.sig.dest, data.ap$dest)[1:length(all.sig.dest)]

all.sig.names <- data.ap$name[sig.dest.rows]

apdelay <- aggregate(data.ap[,6], list(data.ap$dest), mean, na.rm = T)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
as.character(t(all.sig.names))
```

While 30 airports significantly affected the cancelation of flights:
``` {r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
reg.ap <- summary(glm(cancel~dest, data = data.ap))

sig <- 0
j <- 1

for (i in 2:(nrow(coef(reg.ap)))) {
  if (coef(reg.ap)[i,4] < .05) {
    sig[j] <- i
    j <- j + 1
  }
}

all.dest <- sort(unique(data.ap$dest))

all.sig.dest <- all.dest[sig]

sig.dest.rows <- match(all.sig.dest, data.ap$dest)[1:length(all.sig.dest)]

all.sig.names <- data.ap$name[sig.dest.rows]
```

``` {r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
as.character(t(all.sig.names))
```

``` {r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
data.sub <- data.ap %>%
  group_by(dest) %>%
  summarize(
    delaymean = mean(delay, na.rm = TRUE)
  )

data.su <- data.sub[ which(data.sub$dest == all.sig.dest)]
```

``` {r, echo = FALSE,  message = FALSE, warning = FALSE, fig.height = 4, fig.cap = "A histogram mapping the frequency of mean densities, by airport. The airports are indicated by color, but the legend was much too large (101 airports!) to include."}
ggplot(data = data.sub, aes(x = delaymean, y = ..density.., color = dest), 
       main = "Density of Mean Delay Times, Colored by Destination Airport", 
       xlab = "Delay Mean", ylab = "Density") +
  geom_histogram() + guides(fill = FALSE)
```

## Airplane

None of the engine types are significant by themselves, but various models and plane types (also seen as tail numbers) have a significant effect on delay:
```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
summary(glm(delay~type+manufacturer+model, data = data.p))
```

```{r,echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.cap = "Number of engines on the airplane mapped against the length of delays (and the amount of delays) per amount of engines. The color indicates the type of plane."}
ggplot(data = data.p, aes(x = engines, y = delay, colour = type), 
       main = "Number of Engines and Flight Delay Lengths", 
       xlab = "Engines", ylab = "Delay") +
  geom_point( alpha = .2) +
  scale_size(range = c(4,10)) +
  guides(colour = guide_legend(override.aes = list(alpha = 1, size = 6)))
```

## Conclusion

There seems to be an optimal, minimize-delay-probability set of conditions in which to fly: lower temperatures, midnight to 5 A.M. and with a fixed-wing single engine plane. Because delay time and cancelation probabilities are highly correlated, follow the guidelines to avoid takeoff delays to tyr to avoid flight cancelation. Fly away!